Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-22 manuscript.qmd
add both contra evos to figure & rerun pipeline

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.

The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for late T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension. In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy). These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures
.svg){#fig-model-prevalences-with-ipsi}

@@

In essence, this is a summary of the results in @sec-results-risks. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.

>>>>

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-algebra-equation, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in @tbl-midline-params.

The matrix on the left side of @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side show the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for late T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension. In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-algebra-equation}

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy). These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

@@

In essence, this is a summary of the results in @sec-results-risk. However, it should be considered in light of the limitations outlined in @sec-discussion-limitations below.
